### 2021-01-01 FrontBackmatter/Abstract.tex
Finish first draft of chapters 1 and 2

\chapter*{Abstract}
Short summary of the contents in English\dots a great guide by
Kent Beck how to write good abstracts can be found here:

>>>>

\chapter*{Abstract}
Biased data represents a significant problem for the proper functioning of machine learning models,
which affects the trustworthiness of deployed models.
These biases are usually introduced by the data generation process,
i.e., data is collected from non-representative samples or the result of biased processes.
However, these data deficiencies can be very expensive or even impossible to fix,
which makes it desirable to solve the problem on the algorithmic end.
In this work, I consider two different forms of data bias:
labeling bias and sampling bias;
investigated under the framework of algorithmic fairness and evaluated using common fairness metrics.
Labeling bias here refers to a systematic bias,
correlated with a sensitive attribute,
which causes the labels in the dataset to differ from the ``true'' labels;
whereas sampling bias indicates that samples are missing from the training set in a systematic way,
but are still present in the setting where the model is intended to be deployed.
Both biases will make a naively trained model fail to generalize.
I present three approaches to tackling this problem, each relying on some form of additional knowledge about the data.
The first approach, dealing with labeling bias, is based on implicit, probabilistic target labels
which satisfy certain given statistics.
These target labels can be used to train any likelihood-based model.
The second approach deals with strong spurious correlations in the training data,
which can be seen as a specific form of sampling bias.
A bias-free partially-labeled context set is used to learn an interpretable representation of the data
which is invariant to the spurious correlation and can be assessed qualitatively.
The third approach deals with less extreme cases of sampling bias,
but relaxes the assumption of having any labels in the context set
by learning an invariant representation via distribution matching.

### 2021-01-01 ch/1-introduction.tex
Finish first draft of chapters 1 and 2

When evaluating models in these setup,
the problem is often that we do not have access to the underlying true unbiased dataset.
In this case, we cannot just rely on metrics like accuracy to inform us how well the model is doing:
the model is not being evaluated on the right dataset, so accuracy is not that informative.
For these cases where we don't have access to a truly unbiased test set,
we use \emph{fairness metrics}, which do not rely on access to an unbiased set.
These fairness metrics, which are based on different fairness definitions,
are then used alongside metrics like accuracy to evaluate how well a model has corrected for the training set bias.
Note however, that evaluation on a truly unbiased dataset is still preferred:
fairness metrics can make assumptions about what is ideal that might not fit the actual use case.

@@

The fairness definition that most fits the described setup of label bias and sampling bias is \ac{DP},
also called statistical parity or independence.
It demands that the predictions $\hat{y}$ be independent of the sensitive attribute.
So, for binary $s$ and $y$:
\begin{align}
  P(\hat{y}=1|s=0) &= P(\hat{y}=1|s=1)~.
  \label{eq:dp-def}
\end{align}
There are multiple \ac{DP} \emph{metrics} which track how close the predictions are to satisfying the equality,
among them the difference and the ratio of the terms on the two sides of the equation.

When we previously talked about unbiased datasets,
we did not specify exactly what this meant,
and that is because this can differ from task to task,
but one way to define it is as a \emph{balanced dataset} where all combinations of $s$ and $y$ occur at the same rate:
\begin{align}
  \label{eq:balanced-dataset}
  P(y=0,s=0)=P(y=0,s=1)=P(y=1,s=0)=\dots
\end{align}
In such a dataset, we have $y \perp s$, and thus,
perfect predictions ($\hat{y}=y$) on this dataset will satisfy $\hat{y} \perp s$ and hence \ac{DP}.
We can conclude that perfect accuracy on a balanced test set implies \acl{DP} (the opposite does not hold).
However, if a model satisfies \ac{DP} on a \emph{biased} dataset,
then it necessarily cannot be perfectly accurate with respect to that dataset anymore,
which makes sense because the goal is to be accurate to the \emph{unbiased} dataset.
This leads to a fairness-accuracy trade-off on biased test sets.

>>>>

When evaluating models in these setup,
the problem is often that we do not have access to the underlying true unbiased dataset.
In this case, we cannot just rely on metrics like accuracy to compare models:
the model is not being evaluated on the right dataset, so accuracy is not that informative.
For these cases, where we do not have access to a truly unbiased test set,
we use \emph{fairness metrics}, which do not rely on access to an unbiased set.
These fairness metrics, which are based on different fairness definitions,
are then used alongside metrics like accuracy to evaluate how well a model has corrected for the training set bias.
Note however, that evaluation on a truly unbiased dataset is still preferred:
fairness metrics make assumptions about what it means to be unbiased that might not fit the actual use case.

@@

The fairness definition that most fits the described setup of label bias and sampling bias is \acf{DP},
also called statistical parity or independence.
It demands that the predictions $\hat{y}$ be independent of the sensitive attribute \(s\).
So, for binary $s$ and $y$:
\begin{align}
  P(\hat{y}=1|s=0) &= P(\hat{y}=1|s=1)~.
  \label{eq:dp-def}
\end{align}
There are multiple \ac{DP} \emph{metrics} which track how close the predictions are to satisfying the equality,
among them the difference and the ratio of the terms on the two sides of the equation.

When we previously talked about unbiased datasets,
we did not specify exactly what this meant,
and that is because this can differ from task to task,
but one way to define it is as a \emph{balanced} dataset where all combinations of $s$ and $y$ occur at the same rate:
\begin{align}
  \label{eq:balanced-dataset}
  P(y=0,s=0)=P(y=0,s=1)=P(y=1,s=0)=\dots
\end{align}
In such a dataset, we have $y \perp s$, and thus,
perfect predictions ($\hat{y}=y$) on this dataset will satisfy $\hat{y} \perp s$ and hence \ac{DP}.
We can conclude that perfect accuracy on a balanced test set implies \acl{DP} (the opposite does not hold).
However, if a model's predictions satisfy \ac{DP} on a \emph{biased} dataset,
then they cannot be perfectly accurate with respect to that dataset's biased labels anymore,
which makes sense because the goal is to be accurate to the \emph{unbiased} dataset.
This leads to a fairness-accuracy trade-off on biased test sets.

### 2021-01-01 ch/2-related-work.tex
Finish first draft of chapters 1 and 2

\chapter{Related work}\label{ch:related-work}
%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2011fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalization that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
\section{Two views of the dataset bias problem}\label{two-views-of-the-dataset-bias-problem}
As hinted to in the introduction, there is a divide in the related literature in how the problem is defined.
On the one hand, there is the ``groundtruth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is used.

@@

An instance of this setting has been formalized in \citet{blum2020recovering}.
In their formalization, the training set starts out unbiased,
but then, for samples with \(s=0\), class labels \(y\) are flipped from 1 to 0 with probability \(\nu\).
In addition, samples with \(s=0\) and \(y=0\) are dropped with probability \(1-\beta_\mathit{NEG}\)
and those with \(s=0\) and \(y=1\) are dropped with probability \(1-\beta_\mathit{POS}\).

On the other hand, there is the ``definition-centric'' view that given some training data,
and not necessarily knowing anything about the deployment setting,
we define fairness criteria which a fair classifier should satisfy.
Here, the goal is not necessarily to correct for a bias in the training data,
but simply to ensure that the classifier is fair in some specific sense.
By default, a classifier would \emph{not} be fair in that specific way.
It is left open whether this unfairness originates in the training data or in the algorithm itself.

This setting has a fairness-accuracy trade-off.

The definition-centric view was the one used by the first works in the field.
The first fairness definition was Demographic Parity. \textbf{citation}

The difference between these views is not completely clear-cut.
Consider the case where a biased training set is available,
and an unbiased test set is assumed to exist, but we do not have access to it.
Then we might at least know that the unbiased test set satisfies demographic parity.
In this case, testing the classifier for demographic parity is an indication
for how it would fare on the unbiased test set (which satisfies DP).

Furthermore, there is significant overlap in the solutions employed for the two views:
a method developed for one of them can often be used for the other as well.

@@

\subsection{Precursor}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data mining.
The first fundamental conclusion of the authors
is that it is not enough to take out the discriminatory features
(referred to as \emph{sensitive attributes} in other works) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
The experiments were done on the ``German credit dataset''
where the task is to classify people as a good or a bad credit risk.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features.
% The latter case can be
% detected by checking the other rules in the dataset.

@@

Finally, as mentioned in the beginning,
\citet{blum2020recovering} provide a formalization of a setting with label bias and sampling bias
and an unbiased ground truth test set.
This is similar to \citet{wick2019unlocking}.

>>>>

\chapter{Related work}\label{ch:related-work}
%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2011fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalization that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
\section{Two views of the dataset bias problem}\label{two-views-of-the-dataset-bias-problem}
As hinted to in the introduction,
there is a divide in the relevant literature in how the problem of dataset bias is approached.
On the one hand, there is the ``groundtruth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is in use.

@@

An instance of this setting has been formalized by \citet{blum2020recovering}.
In their formalization, the training set starts out unbiased,
but then, for samples with \(s=0\), class labels \(y\) are flipped from 1 to 0 with probability \(\nu\).
In addition, samples with \(s=0\) and \(y=0\) are dropped with probability \(1-\beta_\mathit{NEG}\)
and those with \(s=0\) and \(y=1\) are dropped with probability \(1-\beta_\mathit{POS}\).
Meanwhile, the test remains unbiased and is the basis for evaluating our models.

On the other hand, there is the ``definition-centric'' view that,
given some training data and not necessarily knowing anything about the deployment setting,
we define \emph{fairness criteria} which a fair classifier should satisfy.
Here, the goal is not primarily to correct for the deviation of the training data from the true distribution,
but simply to ensure that the classifier is fair in some specific sense.
By default, a classifier would \emph{not} be fair in that specific way.
It is left open whether this unfairness originates in the training data or in the algorithm itself.
Evaluation is typically performed on test set that has all the same biases as the training set,
and thus a fairness-accuracy trade-off arises.

The first works in the field are almost exclusively took the definition-centric view,
most of them using \acf{DP} \citep{dwork2012fairness} as the fairness criterion.

The difference between these views is not completely clear-cut.
Consider the case where a biased training set is available,
and an unbiased test set is assumed to exist, but we do not have access to it \citep{jiang2020identifying}.
Then we might at least know that the unbiased test set satisfies demographic parity.
In this case, testing the classifier for demographic parity is an indication
for how it would fare on the unbiased test set.

Furthermore, there is significant overlap in the solutions employed for the two views:
a method developed for one of them can often be adapted to the other as well.

@@

\section{Precursors}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data mining.
The first fundamental conclusion of the authors
is that it is not enough to take out the discriminatory features
(referred to as \emph{sensitive attributes} in other works) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
The experiments were done on the ``German credit dataset''
where the task is to classify people as a good or a bad credit risk.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features.
% The latter case can be
% detected by checking the other rules in the dataset.

@@

A conceptually very influential work was \citet{friedler2016possibility}.
While this work does not explicitly present a groundtruth-centric approach to bias,
their concepts of the construct space and the observed space are close
to the idea of the (biased) training distribution and the true underlying distribution.

\citet{jiang2020identifying} also formulate their problem this way:
in their setting, the underlying unbiased labels become corrupted by a biased labeler,
which results in a biased training set.
Their goal is to train a classifier that makes correct predictions consistent with the true labels.
However, they treat the true labels as completely unknown
and evaluate their models w.r.t.\ the common fairness definitions on a test set
that is just as biased as the training set.

Finally, as mentioned in the beginning,
\citet{blum2020recovering} provide a formalization of a setting with label bias and sampling bias
and an unbiased ground truth test set.
They explicitly state not being affected by a fairness-accuracy trade-off,
because the true distribution in the considered problem satisfies \acf{DP}.
Their formalized dataset generation process has multiple steps:
First, a small amount of random noise is applied to the (true) labels;
this models general inaccuracies in the labeling process and does not introduce bias yet.
Second, sampling bias is added, by dropping samples based on what the sample's \(s\) and \(y\) values are.
Concretely, only samples with \(s=0\) are dropped.
Finally, labeling bias is added, by flipping labels for one of the demographic groups (\eg, \(s=0\)),
in one specific direction (from \(y=1\) to \(y=0\)).
The authors show that in this specific situation,
an unbiased classifier (as determined by evaluation on the unbiased test set) can be recovered
by enforcing its predictions to be compatible with \acf{EOpp} on the training set.
Notably, the fairness constraint enforced during training (\ac{EOpp})
is not the fairness constraint satisfied by the predictions on the unbiased test set (\ac{DP}).
One reason, why enforcing \ac{EOpp} works so well here is
that the label bias only affects the \emph{underrepresented} demographic group,
and that the label bias is is the direction (\(y=1\) to \(y=0\)) that \ac{EOpp} is most sensitive to.
Thus, even in the groundtruth-centric view of bias,
fairness constraints can be very useful.
